---
title: "RT-PCR"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

## Global parameters

The data are in this folder:

```{r}
data_folder <- "/Users/MarcChoisy/Library/CloudStorage/OneDrive-OxfordUniversityClinicalResearchUnit/GitHub/choisy/RT-PCR/data"
```

## Packages

Installing the required packages if not already installed:

```{r}
needed <- c("purrr", "stringr", "mixtools", "quantmod", "magrittr", "dplyr")
to_install <- needed[!needed %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install)
```

Loading the packages:

```{r message = FALSE}
for (package in needed) library(package, character.only = TRUE)
```

## Functions

The $x$ and $y$ labels of the plots:

```{r}
xlab <- "temperature (°C)"
ylab <- "(-d/dT) fluorescence (465-510)"
```

A few utilitary functions:

```{r}
lines2 <- function(...) lines(..., lwd = 3)

plotl <- function(...) plot(..., type = "l", lwd = 3)

plot_fluo <- function(...) plot(..., xlab = xlab, ylab = ylab)

plotl_fluo <- function(...) plotl(..., xlab = xlab, ylab = ylab)

findPeaks2 <- function(...) {findPeaks(...) - 1}

legend2 <- function(...) legend("topleft", ..., lty = 1, lwd = 3, bty = "n")

grep_val <- function(...) grep(..., value = TRUE)
```


## Loading the data

Loading the data from disk:

```{r}
# names of the files:
slot_names <- data_folder |> 
  dir() |>
  str_remove(".csv")

# loading all the data into a list of data frames:
rtpcr <- data_folder |>
  dir(full.names = TRUE) |> 
  map(read.csv) |> 
  setNames(slot_names)
```

## Figures

Let's choose one example:

```{r}
a <- rtpcr[["021120"]]
```

It looks like this:

```{r}
plotl_fluo(a[[1]], a[[2]], col = 4)
```

## Finding peaks

The function `findPeaks2()` allows to easily find peaks:

```{r}
ind <- findPeaks2(a[[2]], .05)
plotl_fluo(a[[1]], a[[2]], col = 4)
points(a[[1]][ind], a[[2]][ind], col = 2, lwd = 3)
```

## Finding baseline noise

The function below extracts the baseline noise from the `x` 2-column dataframe
in which the first and second columns correspond to the $x$ and $y$ values
respectively:

```{r}
extract_noise <- function(x, t1, t2, ...) {
  x |> 
    filter(if_any(1, ~ . < t1 | t2 < .)) |> 
    approx(...) |> 
    extract(c("x", "y")) |> 
    as.data.frame()
}
```

Let's try it:

```{r}
noise <- extract_noise(a[, 1:2], 76, 91)
plotl_fluo(a[[1]], a[[2]], col = 4)
lines2(noise, col = 2, lty = 2)
```

## Mixture model

Removing noise:

The `b` data frame is a duplicate of the `a` data frame in which the baseline
noise has been removed:

```{r}
b <- a
b[, 2] <- (b[[2]] - extract_noise(b[, 1:2], 76, 91, xout = b[[1]])$y) |> 
  map_dbl(max, 0)
plotl_fluo(b[[1]], b[[2]], col = 4)
```

The following function disaggregate the -dF/dT and T data into T data:

```{r}
disaggregate <- function(x, n) {
  rep(x[[1]], round(n * x[[2]]))
}
```

Normal mixture model:

```{r}
fmm <- b |> 
  disaggregate(10) |> 
  normalmixEM(1 / length(ind), a[[1]][ind], 1.5)
```

Let's compare:

```{r}
xval <- seq(75, 90, le = 500)

components <- fmm |> 
  extract(c("mu", "sigma", "lambda")) |>
  pmap(~ dnorm(xval, ..1, ..2) * ..3) |> 
  as.data.frame()

yval <- rowSums(components)

eps <- sum(unique(round(diff(b[, 1]), 7)) * b[, 2]) /
  sum(unique(round(diff(xval), 7)) * yval)

plotl_fluo(b[[1]], b[[2]], col = 4)
lines2(xval, eps * yval, col = 2)

legend2(c("raw data", "model"), col = c(4, 2))
```

Let's plot the individual components:

```{r}
plotl_fluo(b[[1]], b[[2]], col = 4)
walk(eps * components, ~ lines2(xval, .x, col = 2))
legend2(c("raw data", "model"), col = c(4, 2))
```

Let's look at the estimated standard deviations of the individual components:

```{r}
fmm$sigma
```

All equal, which suggest that from now we can just look at the peak value as a
good proxy of concentration. But let's compare the raw peak values with the
processed ones. Raw peak values are:

```{r}
(raw_peaks <- a[[2]][ind])
```

After the processing, the peak values are:

```{r}
(pcs_peaks <- unname(map_dbl(eps * components, ~ .x[findPeaks2(.x)])))
```

```{r}
opar <- par(pty = "s")
lims <- c(6, 19)
plot(pcs_peaks, raw_peaks, xlim = lims, ylim = lims, col = 4, lwd = 3, pch = 3,
     xlab = "corrected values", ylab = "raw values")
abline(0, 1)
par(opar)
```

## Looking at all the data

Some utilitary functions:

```{r}
multi_type <- function(x, multi = "^multi1.*") {
  x[c("Tm", grep_val(multi, names(x)))]
}

remove_1col_df <- function(x) {
  x[map_int(x, ncol) > 1]
}
```

The following function draws all the lines of one data frame:

```{r}
plotx <- function(x, y = NA) {
  xval <- x[[1]]
  yval <- x[, -1]
  plot(rep(xval, ncol(yval)), unlist(yval), type = "n", xlab = xlab, ylab = ylab)
  walk(yval, ~ lines(xval, .x, col = 4))
  title(y)
}
```

Let's try it:

```{r}
plotx(a[c("Tm", grep_val("^multi1.*", names(a)))])
plotx(a[c("Tm", grep_val("^multi2.*", names(a)))])
```

The following function draws of the lines of a list of data frames:

```{r}
plotxx <- function(x) {
  f <- function(y) {
    xval <- y[[1]]
    yval <- y[, -1]
    data.frame(rep(xval, ncol(yval)), unlist(yval))
  }
  
  g <- function(y) {
    xval <- y[[1]]
    yval <- y[, -1]
    walk(yval, ~ lines(xval, .x, col = adjustcolor(4, .1)))
  }
  
  x |>
    map(f) |> 
    bind_rows() |> 
    plot(type = "n", xlab = xlab, ylab = ylab)
  
  walk(x, g)
}
```

Let's try it on the `multi1`:

```{r}
rtpcr |>
  map(multi_type, "^multi1.*") |> 
  remove_1col_df() |> 
  plotxx()
```

And on the `multi2`:

```{r}
rtpcr |>
  map(multi_type, "^multi2.*") |> 
  remove_1col_df() |> 
  plotxx()
```

```{r}
multi1 <- rtpcr |>
  map(multi_type, "^multi1.*") |> 
  remove_1col_df()
```



